\documentclass{beamer}
\usepackage{../../shared/styles/custom}
\usepackage{../../shared/styles/conventions}

%\beamerdefaultoverlayspecification{<+->}
% \newcommand{\data}{\mathcal{D}}
% \newcommand\Item[1][]{%
% 	\ifx\relax#1\relax  \item \else \item[#1] \fi
% 	\abovedisplayskip=0pt\abovedisplayshortskip=0pt~\vspace*{-\baselineskip}}

\graphicspath{ {../assets/bias-variance/figures/} }

\title{Bias-Variance and Cross Validation}
\date{\today}
\author{Nipun Batra and teaching staff}
\institute{IIT Gandhinagar}
\begin{document}
\maketitle

\begin{frame}As depth increases, train accuracy improves\\
	\pause As depth increases, test accuracy improves till a point\\
	\pause At very high depths, test accuracy is not good (overfitting). 

	\end{frame}

	\begin{frame}\item What causes high variance in a model? Give an example.
\pause
\item How does cross-validation help in model selection?
\pause
\item Why can't we directly optimize for test error?
\end{enumerate}
\end{frame}

\begin{frame}{Key Takeaways}
\begin{itemize}
\item \textbf{Bias-Variance Decomposition}: Total error = Bias² + Variance + Noise
\pause
\item \textbf{High Bias}: Underfitting, model too simple
\pause
\item \textbf{High Variance}: Overfitting, model too complex
\pause
\item \textbf{Cross-Validation}: Essential for proper model evaluation
\pause
\item \textbf{Model Selection}: Choose complexity that balances bias and variance
\pause
\item \textbf{No Free Lunch}: Cannot reduce both bias and variance simultaneously
\end{itemize}
\end{frame}

\begin{frame}{Next time: Ensemble Learning}
\begin{itemize}
\item How to combine various models?
	\item Why to combine multiple models?
	\pause
\item How can we reduce bias?
	\item How can we reduce variance?
\end{itemize}
\end{frame}

\end{document}
